Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-19 optim.qmd
add abstract

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - projection pursuit
  - optimization
  - jellyfish optimiser
  - data visualisation
  - high-dimensional data
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
---

@@

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#install the latest version of ferrn from: https://github.com/huizezhang-sherry/ferrn
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *projection
pursuit index (PPI)*, (@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathcal{R}^{n }\times p$ and PPI
\$f(\\cdot)\$, PP finds the orthonormal projection matrix
$A \in \mathcal{R}^{p \times d}$ by solving the following optimisation
problem:

>>>>

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  Projection pursuit (PP) is a dimension reduction technique that identifies low-dimensional projections in high-dimensional data by optimising a criteria function known as the PP index. The optimisation in PP can be non-smooth and requires identifying optima with a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recent swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated with JSO hyper-parameters to determine their effect on JSO success rate. The JSO algorithm has been implemented in the `tourr` package, while calculations for smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - Jellyfish Search Optimiser (JSO)
  - Optimisation
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
---

@@

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
#install the latest version of ferrn from: https://github.com/huizezhang-sherry/ferrn
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *projection
pursuit index (PPI)*, (@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathcal{R}^{n }\times p$ and PPI
$f(\cdot)$, PP finds the orthonormal projection matrix
$A \in \mathcal{R}^{p \times d}$ by solving the following optimisation
problem:

### 2024-06-19 optim.tex
add abstract

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Leung},
  pdfkeywords={projection pursuit, optimization, jellyfish
optimiser, data visualisation, high-dimensional data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Studying the Performance of the Jellyfish Search Optimiser for
the Application of Projection Pursuit}
\author[1]{H. Sherry Zhang%
\corref{cor1}%
}
 \ead{huize.zhang@austin.utexas.edu} 
\author[2]{Dianne Cook%
%
}
 \ead{dicook@monash.edu} 
\author[3]{Nicolas Langrené%
%
}
 \ead{nicolaslangrene@uic.edu.cn} 
\author[2]{Jessica Leung%
%
}
 \ead{Jessica.Leung@monash.edu}

@@

\begin{abstract}
This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
tempus elit.
\end{abstract}

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}

@@

The artificial jellyfish search (JS) algorithm \citep{chou_novel_2021}
is a swarm-based metaheuristic optimisation algorithm inspired by the
search behaviour of jellyfish in the ocean. It is one of the newest
swarm intelligence algorithms \citep{rajwar_exhaustive_2023}, which was
shown to have stronger search ability and faster convergence with few
algorithmic parameters compared to classic optimization methods
\citep{chou_novel_2021}-\citep{chou_recent_2022}.

Effective optimisation is an important aspect of many methods employed
for visualising high-dimensional data (\(X\)). Here we are concerned
about computing informative linear projections of high-dimensional
(\(p\)) data using projection pursuit (PP) (\citet{kr69}, \citet{FT74}).
This involves optimising a function (e.g. \citet{hall1989polynomial},
\citet{cook1993projection}, \citet{lee2010projection},
\citet{Loperfido2018}, \citet{Loperfido2020}), called the projection
pursuit index (PPI), that defines what is interesting or informative in
a projection.

These PPI are defined on projections (\(XA\)), which means that there is
a constraint that needs to be considered when optimising. A projection
of data is defined by a \(p\times d\) orthonormal matrix \(A\), and this
imposes the constraint on the elements of \(A\), that columns need have
norm equal to 1 and the product of columns need to sum to zero.

\citet{cook1995grand} introduced the PP guided tour, which enabled
interactive visualisation of the optimisation in order to visually
explore high-dimensional data. It is implemented in the R \citep{R}
package \texttt{tourr} \citep{tourr}. The optimisation that is
implemented is fairly basic, and potential problems were highlighted by
\citet{RJ-2021-105}. Implementing better optimisation functionality is a
goal, but it needs to be kept in mind that the guided tour also has
places importance on watching the projected data as the optimisation
progresses.

Here we explore the potential for a jellyfish optimisation to be
integrated with the guided tour. Section~\ref{sec-background} explains
the optimisation that is used in the current the projection pursuit
guided tour. Section~\ref{sec-theory} provides more details on the
jellyfish optimiser and formalises several characteristics of projection
pursuit indexes that are help to measure optimiser performance.
\textbf{?@sec-simulation} describes a simulation study on performance of
the jellyfish for several types of data and index functions.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, left=2mm, breakable, toprule=.15mm, colframe=quarto-callout-note-color-frame, opacityback=0, titlerule=0mm, bottomtitle=1mm, colback=white, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, coltitle=black]

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, Jellyfish Search Optimiser
(JSO), Optimisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Studying the Performance of the Jellyfish Search Optimiser for
the Application of Projection Pursuit}
\author[1]{H. Sherry Zhang%
\corref{cor1}%
}
 \ead{huize.zhang@austin.utexas.edu} 
\author[2]{Dianne Cook%
%
}
 \ead{dicook@monash.edu} 
\author[3]{Nicolas Langrené%
%
}
 \ead{nicolaslangrene@uic.edu.cn} 
\author[2]{Jessica Wai Yin Leung%
%
}
 \ead{Jessica.Leung@monash.edu}

@@

\begin{abstract}
Projection pursuit (PP) is a dimension reduction technique that
identifies low-dimensional projections in high-dimensional data by
optimising a criteria function known as the PP index. The optimisation
in PP can be non-smooth and requires identifying optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recent
swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising
PP indexes. The performance of JSO is evaluated across various
hyper-parameter settings and compared with existing optimisers.
Additionally, this work proposes novel methods to quantify two
properties of the PP index -- smoothness and squintability -- that
capture the complexities inherent in PP optimisation problems. These two
metrics are evaluated with JSO hyper-parameters to determine their
effect on JSO success rate. The JSO algorithm has been implemented in
the \texttt{tourr} package, while calculations for smoothness and
squintability are available in the \texttt{ferrn} package.
\end{abstract}

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely \emph{projection
pursuit index (PPI)}, (\citet{hall1989polynomial},
\citet{cook1993projection}, \citet{lee2010projection},
\citet{Loperfido2018}, \citet{Loperfido2020}). Given high-dimensional
data \(X \in \mathcal{R}^{n }\times p\) and PPI \(f(\cdot)\), PP finds
the orthonormal projection matrix \(A \in \mathcal{R}^{p \times d}\) by
solving the following optimisation problem:

\[
\underset{A}{\max} f(XA) \quad \text{subject to} \quad A'A = I
\]

These indices \(f(\cdot)\) are characterised by the informativeness or
``interestingness'' of a projection and are often non-linear and
non-convex. As such, an effective and efficient optimisation procedure
is essential to enable sufficient exploration of the data landscape
during the visualisation process and to arrive at a global optimal
viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP and facilitating a better experience in the visualisation process.
\citet{cook1995grand} introduced the PP guided tour, which enabled
interactive visualisation of the optimisation to visually explore
high-dimensional data. It is implemented in the R \citep{R} package
\texttt{tourr}\citep{tourr}. \citet{RJ-2021-105} highlighted potential
problems of the optimisation routine implemented. While improving the
quality of the optimisation solutions in the tour is essential, it is
also important to be able to watch the projected data as the
optimisation progresses. Therefore, it is of interest to integrate the
guided tour with a global optimisation algorithm and examine the
characteristics and advantages of such implementation.

The artificial jellyfish search (JS) algorithm \citep{chou_novel_2021}
is a swarm-based metaheuristic optimisation algorithm inspired by the
search behaviour of jellyfish in the ocean. It is one of the latest
swarm intelligence algorithms \citep{rajwar_exhaustive_2023}, which was
shown to have stronger search ability and faster convergence with fewer
parameters compared to classic optimisation methods
\citep{chou_novel_2021}-\citep{chou_recent_2022}.

The primary goal of this study is to investigate the performance of JSO
in the context of the PP guided tour. A series of simulation experiments
using well-recognised data sets and PPIs are conducted to yield insights
regarding the behaviour of JSO under different settings and the
sensitivity of hyper-parameter in the optimisation. Second, to observe
the performance of JSO with different types of PPI, a collection of
metrics is introduced to capture specific properties of the index
including squintability and smoothness as introduced in
\citet{laa_using_2020}. To the best of the author's knowledge, this is
the first attempt to quantitatively measure squintability and
smoothness. Finally, the relationship between the JSO performance,
hyper-parameters tuning and various properties of PPI is analysed to
provide helpful guidance for practitioners that are using the guided
tour for high dimensional visualisation.

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the PP guided tour.
Section~\ref{sec-theory} describes the details of JSO and introduces
metrics that measure different characteristics of a PPI.
\textbf{?@sec-simulation} describes a simulation study on the
performance of the JSO using well-known data sets and index functions.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toptitle=1mm, opacityback=0, toprule=.15mm, leftrule=.75mm, coltitle=black, bottomtitle=1mm, arc=.35mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, left=2mm, breakable]
